---
title: "Data Viz: Exercises"
author: "Kylie Ariel Bemis"
date: "11/8/2020"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Grammar of Graphics with ggplot2

## Exercise A1

Create boxplots of Log2Intensity vs. Subject, using the color of the boxplots to represent TechRep.

```{r eval=FALSE}
ggplot(data=iprg,
       mapping=aes(x=Subject, y=Log2Intensity,
                   fill=TechRep)) +
  geom_boxplot()
```

## Exercise A2

Modify your plot from A1 to also facet by TechRep. Encoding the same variable in multiple ways can enhance ease of interpretation.

```{r eval=FALSE}
ggplot(data=iprg,
       mapping=aes(x=Subject, y=Log2Intensity,
                   fill=TechRep)) +
  geom_boxplot() +
  facet_wrap(~TechRep)
```

## Exercise A3

Create boxplots of Intensity vs. Run, using the color of the boxplots to represent Subject. Set the scale of the y-axis to use a logarithmic scale.

```{r eval=FALSE}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Intensity,
                   fill=Subject)) +
  geom_boxplot() +
  scale_y_log10()
```

## Exercise A4

Overlay violin plots and a jittered scatter plot of Log2Intensity vs. Run, using the color of the boxplots to represent Subject.

```{r eval=FALSE}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Log2Intensity,
                   fill=Subject)) +
  geom_jitter(alpha=0.1) +
  geom_violin()
```

## Exercise A5

Plot histograms of Log2Intensity, faceted by Subject and TechRep. Use the color of the histograms to indicate Subject.

```{r eval=FALSE}
ggplot(iprg, aes(x=Log2Intensity, fill=Subject)) +
  geom_histogram() +
  facet_grid(Subject~TechRep)
```

# Statistical graphics for MS

## Exercise B1

Cut the dendrogram for the CRC data to 4 clusters.

Then create a scatter plot of the first two proteins' intensities.

Use the color of the points to represent the clusters, and the shape of the points to represent diagnosis.

```{r eval=FALSE}
plot(crc, col=cutree(hc, k=4), pch=diagnosis)
```

## Exercise B2

Use the k-means clustering for the CRC data with 2 clusters to do the following.

Create a scatter plot of the intensities for AIAG-Bovine and FETUA-Bovine.

Use the color of the points to represent the clusters, and the shape of the points to represent diagnosis.

```{r eval=FALSE}
plot(crc[,c("AIAG-Bovine", "FETUA-Bovine")], col=km$cluster, pch=diagnosis)
```

## Exercise B3

Use the PCA for the CRC data to do the following.

Create a scatter plot of the scores of PC2 and PC3.

Use the color of the points to represent diagnosis.

```{r eval=FALSE}
plot(pc$x[,2:3], col=factor(diagnosis))
```

## Exercise B4

Use the p-values from the models fit for iPRG to create a volcano plot.

Color the points to indicate p-values significant at the alpha = 0.10 level.

Draw dotted lines to show the alpha = 0.10 cutoff and fold changes of 1 and -1.

```{r eval=FALSE}
tests %>%
  mutate(Significance=ifelse(adj.pvalue < 0.1, "P < .10", "P ≥ .10")) %>%
  ggplot(aes(x=log2FC, y=-log10(adj.pvalue))) +
  geom_point(aes(color=Significance)) +
  geom_hline(yintercept=-log10(0.1), linetype="dotted") +
  geom_vline(xintercept=c(-1, 1), linetype="dotted")
```

## Exercise B5

Add axis labels and a title to your plot from B4.

Then export it to a 750 x 500 PNG file.

```{r eval=FALSE}
tiff("ex-figure.tiff", width=750, height=500)
tests %>%
  mutate(Significance=ifelse(adj.pvalue < 0.1, "P < .10", "P ≥ .10")) %>%
  ggplot(aes(x=log2FC, y=-log10(adj.pvalue))) +
  geom_point(aes(color=Significance)) +
  geom_hline(yintercept=-log10(0.1), linetype="dotted") +
  geom_vline(xintercept=c(-1, 1), linetype="dotted") +
  labs(x=expression(log[2]~fold-change),
       y=expression(-log[10]~adjusted~p-value),
       title="Volcano plot")
dev.off()
```



