---
title: "Data Viz: Grammar of Graphics with ggplot2"
author: "Kylie Ariel Bemis"
date: "11/8/2020"
output:
  html_document:
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Why we look at data

Here's a dinosaur dataset:

```{r setup-2, message=FALSE}
library(tidyverse)
library(datasauRus)

filter(datasaurus_dozen, dataset == "dino")
```

If we visualize it, it looks like a T-rex:

```{r dino-plot, echo=FALSE}
datasaurus_dozen %>%
  filter(dataset == "dino") %>%
  ggplot(aes(x=x, y=y)) +
  geom_point() +
  labs(x="x", y="y") +
  theme_minimal()
```

Here's 12 more datasets with identical summary statistics:

```{r dino-dozen-stats, message=FALSE, echo=FALSE}
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarize(
    mean_x    = mean(x),
    mean_y    = mean(y),
    std_dev_x = sd(x),
    std_dev_y = sd(y),
    corr_x_y  = cor(x, y))
```

Based on the summary statistics, we may believe they're similar to each other.

But plotting them reveals quite a difference!

```{r dino-dozen-plots, echo=FALSE}
datasaurus_dozen %>%
  filter(dataset != "dino") %>%
  ggplot(aes(x=x, y=y, colour=dataset)) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(~dataset, ncol=3)
```

Looking at data is important!

- Summary statistics don’t tell whole story

- We can easily find patterns visually

- Identify potential outliers

- Verify model assumptions

- Intuitively display statistical results

# Example data: iPRG

To practice visualization in R, we will use the following iPRG dataset.

The iPRG data is a spiked-in exeriment where 6 proteins were spiked at different ratios in a Yeast proteome background. Each run was repeated three times in a randomized order. The goal is to identify the differentially abundant spiked-in proteins.

>Choi M, et al. ABRF Proteome Informatics Research Group (iPRG) 2015 Study: Detection of Differentially Abundant Proteins in Label-Free Quantitative LC-MS/MS Experiments. J Proteome Res. 2017 Feb 3;16(2):945-957. doi: 10.1021/acs.jproteome.6b00881 Epub 2017 Jan 3. PMID: 27990823.

First, we read in the raw data and annotations.

```{r}
skydir <- "iPRG2015-Skyline"

raw <- read_csv(file.path(skydir, "Choi2017_DDA_Skyline_input.csv"), na="#N/A")
ann <- read_csv(file.path(skydir, "Choi2017_DDA_Skyline_annotation.csv"))
```

Then, we use MSstats to process the raw data.

```{r eval=FALSE}
library(MSstats)

quant <- SkylinetoMSstatsFormat(raw, annotation = ann,
                                removeProtein_with1Feature = TRUE)
quant <- as_tibble(quant)
quant

processed.quant <- dataProcess(quant,
                               normalization="equalizeMedians",
                               summaryMethod="TMP",
                               cutoffCensored="minFeature",
                               censoredInt="0",
                               MBimpute=TRUE,
                               maxQuantileforCensored=0.999)

saveRDS(processed.quant, file="processed-iprg.rds")
```

Having used `saveRDS()` to save the data to "processed-iprg.rds", we can now load the processed data any time we want with `readRDS()`.

We do exactly that below, and then extract the run-level data.

```{r}
processed.quant <- readRDS("processed-iprg.rds")

iprg <- as_tibble(processed.quant$RunlevelData)

iprg
```

We further simplify the dataset to include only relevant columns.

```{r}
iprg <- transmute(iprg,
                  Protein,
                  Run=originalRUN,
                  Log2Intensity=LogIntensities,
                  Intensity=2^LogIntensities,
                  Condition=GROUP_ORIGINAL,
                  Subject=SUBJECT_ORIGINAL,
                  TechRep=stringr::str_sub(originalRUN, -5, -5))
iprg
```

The resulting data frame has 36,321 rows and 5 columns.

# Basics of ggplot2

## Grammar of graphics

The `ggplot2` package implements a *grammar of graphics*.

The idea is that we can build up a plot from its graphical components.

## Components of a plot

The components of the grammar of graphics include:

- A default *tidy* __dataset__ and set of __mappings__ from variables to __aesthetics__

- One or more layers, each having:

    - (Optional) A new dataset
    
    - (Optional) A new set of aesthetic mappings

    - A __geometric object__
    
    - A statistical transformation
    
    - A position adjustment

- A __scale__ for each mapped aesthetic

- A __coordinate system__

- A __facet__ specification, i.e., subplots based on subsets of the data

## Building a basic plot

Suppose we want to visualize the distribution of log-intensities for each run.

We could do this using side-by-side box-and-whisker plots.

### Default dataset

In `ggplot2`, we initiate a plot with a call to the `ggplot()` function.

This sets up the plot's __default dataset__ and a blank canvas.

```{r}
ggplot(data=iprg)
```

### Aesthetic mappings

Next, we add __aesthetic mappings__ with the `aes()` function.

Plot aesthetics include visual channels like position (e.g., the x- and y-axis), color, fill, shape, size, linetype, etc.

Let's map the run and log-intensities to the x- and y-axes.

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Log2Intensity))
```

### Geoms

Now, we need to add marks, or __geometric objects__, to draw the plot.

We can add `geom_boxplot()` to the plot with the `+` operator.

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Log2Intensity)) +
  geom_boxplot()
```

You can find documentation for all `ggplot2` functions here: https://ggplot2.tidyverse.org/reference/

# Building a more complex plot

## More aesthetic mappings

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Log2Intensity,
                   fill=Condition)) +
  geom_boxplot()
```

## Inline transformations

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_boxplot()
```

## Annotations

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run")
```

## Themes

### Preset themes

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal()
```

### Adjusting individual elements

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

## Different geoms

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_violin() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

## Adding layers

### Layering geoms

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_point(position="jitter") +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

### Setting layer aesthetics

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity),
                   fill=Condition)) +
  geom_jitter(size=0.1, alpha=0.1) +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

### Inheriting aesthetic mappings

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=log2(Intensity))) +
  geom_jitter(aes(color=Condition), size=0.1, alpha=0.1) +
  geom_boxplot(aes(fill=Condition)) +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

## Faceting

```{r}
ggplot(data=iprg,
       mapping=aes(x=TechRep, y=log2(Intensity),
                   fill=Condition)) +
  geom_boxplot() +
  labs(x="Technical Replicate", y="Log2 Intensity",
       title="Distribution of log2-intensity by Tech Rep + Condition") +
  facet_grid(~Condition)
```

# Recipes for common plots

## Scatter plot

```{r}
iprg2 <- iprg %>%
  select(Protein, Run, Intensity) %>%
  pivot_wider(names_from=Run, values_from=Intensity)
iprg2
```

```{r}
ggplot(iprg2, aes(x=JD_06232014_sample1_B.raw,
                  y=JD_06232014_sample1_C.raw)) +
  geom_point(alpha=0.1) +
  scale_x_log10() +
  scale_y_log10()
```

## Line plot

```{r}
set.seed(1)
msraw <- as_tibble(Cardinal::simulateSpectrum())
msraw
```

```{r}
ggplot(msraw, aes(x=mz, y=intensity)) +
  geom_line()
```

## Box plot

```{r}
ggplot(iprg, aes(x=Run, y=Log2Intensity)) +
  geom_boxplot() +
  coord_flip()
```

## Histogram

```{r}
ggplot(iprg, aes(x=Log2Intensity)) +
  geom_histogram() +
  facet_wrap(~Run)
```

## Bar plot

```{r}
ggplot(raw, aes(x=ProductCharge)) +
  geom_bar()
```

# Choosing color palettes

## Color brewer

```{r brewer, fig.height=8, fig.width=4}
RColorBrewer::display.brewer.all()
```

## Qualitative palettes

```{r}
ggplot(data=iprg,
       mapping=aes(x=Run, y=Log2Intensity,
                   fill=Condition)) +
  geom_boxplot() +
  labs(x="MS run", y="Log2 Intensity",
       title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_brewer(palette="Dark2")
```

## Continuous palettes

```{r}
volcano2 <- as_tibble(cbind(expand.grid(x=seq_len(nrow(volcano)),
                                        y=seq_len(ncol(volcano))),
                            tibble(z=as.vector(volcano))))
volcano2
```

### Sequential

```{r}
ggplot(volcano2, aes(x=x, y=y, fill=z)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_distiller(palette="Blues")
```

### Diverging

```{r}
volcano2 %>%
  mutate(z0=(z - mean(z)) / sd(z)) %>%
  ggplot(aes(x=x, y=y, fill=z0)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_distiller(palette="PRGn")
```

# Saving figures

## Iterating a plot

```{r}
g <- ggplot(data=iprg,
            mapping=aes(x=Run, y=log2(Intensity)))
```

```{r}
g + geom_jitter(aes(color=Condition), size=0.1, alpha=0.1)
```

```{r}
g2 <- g + geom_jitter(aes(color=Condition),
                      size=0.1, alpha=0.1) +
  geom_boxplot(aes(fill=Condition))

g2
```

## Exporting a figure

R displays plots on "graphic devices".

The best way to export a graphic from R is to create an external graphics device.

You can create external graphics devices for common file formats:

- `pdf()`

- `png()`

- `tiff()`

- `jpeg()`

Here, we export a figure to a PDF device.

```{r}
pdf("test-figure.pdf", height=5, width=10)
g2 + labs(x="MS run", y="Log2 Intensity",
          title="Distribution of log2-intensity by MS run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
dev.off()
```

After plotting the figure, make sure to call `dev.off()` to turn off the device and close the file.

# Session info

```{r session-info}
sessionInfo()
```
